# Getting acquainted with R {#exercise1}

```{r "setup", include = FALSE}
require("knitr")
opts_knit$set(root.dir = "data/")
```

In this exercise we will use `R` to read a dataset and produce some descriptive statistics, produce some charts, and perform some simple statistical inference. The aim of the exercise is for you to become familiar with `R` and some basic `R` functions and objects.

The first thing we will do, after starting `R`, is issue a command to retrieve an example dataset: 

```{r, eval = TRUE}
fem <- read.table("fem.dat", header = TRUE)
```

This command illustrates some key things about the way `R` works.

We are instructing `R` to assign (using the `<-` operator) the output of the `read.table()` function to an object
called `fem`.

The `fem` object will contain the data held in the file `fem.dat` as an `R` data.frame object:

```{r, eval = TRUE}
class(fem)
```

You can inspect the contents of the `fem` data.frame (or any other `R` object) just by typing its name: 

```{r, eval = FALSE}
fem
```

```{r, echo = FALSE, eval = TRUE}
head(fem)
```

Note that the `fem` object is built from other objects. These are the named vectors (columns) in the dataset: 

```{r, eval = TRUE}
names(fem)
```

The `[1]` displayed before the column names refers to the numbered position of the first name in the output. These positions are known as indexes and can be used to refer to individual items. For example:

```{r, eval = TRUE}
names(fem)[1]
```

```{r, eval = TRUE}
names(fem)[8]
```

```{r, eval = TRUE}
names(fem)[2:4]
```

The data consist of 118 records:

```{r, eval = TRUE}
nrow(fem)
```

each with nine variables:

```{r, eval = TRUE}
ncol(fem)
```

for female psychiatric patients.

The columns in the dataset are:

+----------+----------------------------------------------------+
| **ID**   | Patient ID                                         |
+==========+====================================================+
| **AGE**  | Age in years                                       |
+----------+----------------------------------------------------+
| **IQ**   | IQ score                                           |
+----------+----------------------------------------------------+
| **ANX**  | Anxiety (1=none, 2=mild, 3=moderate, 4=severe)     |
+----------+----------------------------------------------------+
| **DEP**  | Depression (1=none, 2=mild, 3=moderate or severe)  |
+----------+----------------------------------------------------+
| **SLP**  | Sleeping normally (1=yes, 2=no)                    |
+----------+----------------------------------------------------+
| **SEX**  | Lost interest in sex (1=yes, 2=no)                 |
+----------+----------------------------------------------------+
| **LIFE** | Considered suicide (1=yes, 2=no)                   |
+----------+----------------------------------------------------+
| **WT**   | Weight change (kg) in previous 6 months            |
+----------+----------------------------------------------------+

The first ten records of the `fem` data.frame are:

```{r, echo = FALSE, eval = TRUE}
fem[1:10, ]
```

You may check this by asking `R` to display all columns of the first ten records in the `fem` data.frame:

```{r, eval = TRUE}
fem[1:10, ]
```

The space after the comma is optional. You can think of it as a *placeholder* for where you would specify the indexes for columns you wanted to display. For example:

```{r, eval = FALSE}
fem[1:10,2:4]
```

displays the first ten rows and the second, third and fourth columns of the `fem` data.frame:

```{r echo = FALSE, eval = TRUE}
fem[1:10,2:4 ]
```

`NA` is a special value meaning *not available* or *missing*.

You can access the contents of a single column by name:

```{r, eval = TRUE}
fem$IQ
```

```{r, eval = TRUE}
fem$IQ[1:10]
```

The `$` sign is used to separate the name of the data.frame and the name of the column of interest. Note that `R` is case-sensitive so that `IQ` and `iq` are **_not_** the same.

You can also access rows, columns, and individual cells by specifying row and column positions. For example, the `IQ` column is the third column in the `fem` data.frame:

```{r, eval = TRUE}
fem[ ,3]
```

```{r, eval = TRUE}
fem[9, ]
```

```{r, eval = TRUE}
fem[9,3]
```

There are missing values in the `IQ` column which are all coded as **-99**. Before proceeding we must set these to
the special `NA` value:

```{r, eval = TRUE}
fem$IQ[fem$IQ == -99] <- NA
```

The term inside the square brackets is also an index. This type of index is used to refer to subsets of data held in an object that meet a particular condition. In this case we are instructing `R` to set the contents of the `IQ` variable to `NA` if the contents of the `IQ` variable is **-99**.

Check that this has worked:

```{r, eval = TRUE}
fem$IQ
```

We can now compare the groups who have and have not considered suicide. For example:

```{r, eval = FALSE}
by(fem$IQ, fem$LIFE, summary)
```

Look at the help for the `by()` function:

```{r, eval = FALSE}
help(by)
```

Note that you may use `?by` as a shortcut for `help(by)`.

The `by()` function applies another function (in this case the `summary()` function) to a column in a
data.frame (in this case `fem$IQ`) split by the value of another variable (in this case `fem$LIFE`).

It can be tedious to always have to specify a data.frame each time we want to use a particular variable. We can
fix this problem by ‘attaching’ the data.frame:

```{r, eval = TRUE}
attach(fem)
```

We can now refer to the columns in the `fem` data.frame without having to specify the name of the data.frame. This time we will produce summary statistics for `WT` by `LIFE`:

```{r, eval = TRUE}
by(WT, LIFE, summary)
```

We can view the same data as a box and whisker plot:

```{r, eval = TRUE}
boxplot(WT ~ LIFE)
```

We can add axis labels and a title to the graph:

```{r, eval = TRUE}
boxplot(WT ~ LIFE,
        xlab = "Life",
        ylab = "Weight",
        main = "Weight BY Life")
```

A more descriptive title might be "Weight Change BY Considered Suicide".

The groups do not seem to differ much in their medians and the distributions appear to be reasonably symmetrical about their medians with a similar spread of values.

We can look at the distribution as histograms:

```{r, eval = TRUE}
hist(WT[LIFE == 1])
```

```{r, eval = TRUE}
hist(WT[LIFE == 2])
```

and check the assumption of normality using quantile-quantile plots:
  
```{r, eval = TRUE}
qqnorm(WT[LIFE == 1])
qqline(WT[LIFE == 1])
```

```{r, eval = TRUE}
qqnorm(WT[LIFE == 2])
qqline(WT[LIFE == 2])
```

or by using a formal test:

```{r, eval = TRUE}
shapiro.test(WT[LIFE == 1])
```

```{r, eval = TRUE}
shapiro.test(WT[LIFE == 2])
```

Remember that we can use the `by()` function to apply a function to a data.frame, including statistical functions such as `shapiro.test()`:

```{r, eval - TRUE}
by(WT, LIFE, shapiro.test)
```

We can also test whether the variances differ significantly using *Bartlett’s test* for the homogeneity of variances: 

```{r, eval = TRUE}
bartlett.test(WT, LIFE)
```

There is no significant difference between the two variances.

Many functions in `R` have a *formula interface* that may be used to specify multiple variables and the relations between multiple variables. We could have used the formula interface with the `bartlett.test()` function: 

```{r, eval = TRUE}
bartlett.test(WT ~ LIFE)
```

Having checked the normality and homogeneity of variance assumptions we can proceed to carry out a `t-test`: 

```{r, eval = TRUE}
t.test(WT ~ LIFE, var.equal = TRUE)
```

There is no evidence that the two groups differ in weight change in the previous six months.

We could still have performed a `t-test` if the variances were not homogenous by setting the **var.equal** parameter of the `t.test()` function to **FALSE**: 

```{r, eval = TRUE}
t.test(WT ~ LIFE, var.equal = FALSE)
```

or performed a non-parametric test:

```{r, eval = TRUE}
wilcox.test(WT ~ LIFE)
```

An alternative, and more general, non-parametric test is:

```{r, eval = TRUE}
kruskal.test(WT ~ LIFE)
```

