# Manipulating objects and creating new functions {#exercise2}

In this exercise we will explore how to manipulate `R` objects and how to write functions that can manipulate and extract data and information from `R` objects and produce useful analyses.

Before we go any further we should start `R` and retrieve a dataset:

```{r, eval = TRUE}
salex <- read.table("salex.dat", header = TRUE, na.strings = "9")
```

Missing values are coded as 9 throughout this dataset so we can use the `na.strings` parameter of the `read.table()` function to replace all 9’s with the special `NA` code when we retrieve the dataset. Check that this works by examining the `salex` data.frame:

```{r, eval = TRUE}
salex
names(salex)
```

This data comes from a food-borne outbreak. On Saturday 17th October 1992, eighty-two people attended a buffet meal at a sports club. Within fourteen to twenty-four hours, fifty-one of the participants developed diarrhoea, with nausea, vomiting, abdominal pain and fever.

The columns in the dataset are as follows:

+---------------+----------------------+
| **ILL**       | Ill or not-ill       |
+---------------+----------------------+
| **HAM**       | Baked ham            |
+---------------+----------------------+
| **BEEF**      | Roast beef           |
+---------------+----------------------+
| **EGGS**      | Eggs                 |
+---------------+----------------------+
| **MUSHROOM**  | Mushroom flan        |
+---------------+----------------------+
| **PEPPER**    | Pepper flan          |
+---------------+----------------------+
| **PORKPIE**   | Pork pie             |
+---------------+----------------------+
| **PASTA**     | Pasta salad          |
+---------------+----------------------+
| **RICE**      | Rice salad           |
+---------------+----------------------+
| **LETTUCE**   | Lettuce              |
+---------------+----------------------+
| **TOMATO**    | Tomato salad         |
+---------------+----------------------+
| **COLESLAW**  | Coleslaw             |
+---------------+----------------------+
| **CRISPS**    | Crisps               |
+---------------+----------------------+
| **PEACHCAKE** | Peach cake           |
+---------------+----------------------+
| **CHOCOLATE** | Chocolate cake       |
+---------------+----------------------+
| **FRUIT**     | Tropical fruit salad |
+---------------+----------------------+
| **TRIFLE**    | Trifle               |
+---------------+----------------------+
| **ALMONDS**   | Almonds              |
+---------------+----------------------+

Data is available for seventy-seven of the eighty-two people who attended the sports club buffet. All of the variables are coded 1=yes, 2=no.

We can use the `attach()` function to make it easier to access our data: 

```{r, eval = TRUE}
attach(salex)
```

The two-by-two table is a basic epidemiological tool. In analysing data from a food-borne outbreak collected as a retrospective cohort study, for example, we would tabulate each exposure (suspect foodstuffs) against the outcome (illness) and calculate risk ratios and confidence intervals. `R` has no explicit function to calculate risk ratios from two-by-two tables but we can easily write one ourselves.

