# Manipulating objects and creating new functions {#exercise2}

In this exercise we will explore how to manipulate `R` objects and how to write functions that can manipulate and extract data and information from `R` objects and produce useful analyses.

Before we go any further we should start `R` and retrieve a dataset:

```{r, eval = TRUE}
salex <- read.table("salex.dat", header = TRUE, na.strings = "9")
```

Missing values are coded as 9 throughout this dataset so we can use the `na.strings` parameter of the `read.table()` function to replace all 9’s with the special `NA` code when we retrieve the dataset. Check that this works by examining the `salex` data.frame:

```{r, eval = TRUE}
salex
names(salex)
```

This data comes from a food-borne outbreak. On Saturday 17th October 1992, eighty-two people attended a buffet meal at a sports club. Within fourteen to twenty-four hours, fifty-one of the participants developed diarrhoea, with nausea, vomiting, abdominal pain and fever.

The columns in the dataset are as follows:

+---------------+----------------------+
| **ILL**       | Ill or not-ill       |
+---------------+----------------------+
| **HAM**       | Baked ham            |
+---------------+----------------------+
| **BEEF**      | Roast beef           |
+---------------+----------------------+
| **EGGS**      | Eggs                 |
+---------------+----------------------+
| **MUSHROOM**  | Mushroom flan        |
+---------------+----------------------+
| **PEPPER**    | Pepper flan          |
+---------------+----------------------+
| **PORKPIE**   | Pork pie             |
+---------------+----------------------+
| **PASTA**     | Pasta salad          |
+---------------+----------------------+
| **RICE**      | Rice salad           |
+---------------+----------------------+
| **LETTUCE**   | Lettuce              |
+---------------+----------------------+
| **TOMATO**    | Tomato salad         |
+---------------+----------------------+
| **COLESLAW**  | Coleslaw             |
+---------------+----------------------+
| **CRISPS**    | Crisps               |
+---------------+----------------------+
| **PEACHCAKE** | Peach cake           |
+---------------+----------------------+
| **CHOCOLATE** | Chocolate cake       |
+---------------+----------------------+
| **FRUIT**     | Tropical fruit salad |
+---------------+----------------------+
| **TRIFLE**    | Trifle               |
+---------------+----------------------+
| **ALMONDS**   | Almonds              |
+---------------+----------------------+

Data is available for seventy-seven of the eighty-two people who attended the sports club buffet. All of the variables are coded 1=yes, 2=no.

We can use the `attach()` function to make it easier to access our data: 

```{r, eval = TRUE}
attach(salex)
```

The two-by-two table is a basic epidemiological tool. In analysing data from a food-borne outbreak collected as a retrospective cohort study, for example, we would tabulate each exposure (suspect foodstuffs) against the outcome (illness) and calculate risk ratios and confidence intervals. `R` has no explicit function to calculate risk ratios from two-by-two tables but we can easily write one ourselves.

The first step in writing such a function would be to create the two-by-two table. This can be done with the `table()` function. We will use a table of `HAM` by `ILL` as an illustration:

```{r, eval = FALSE}
table(HAM, ILL)
```

This command produces the following output:

```{r, echo = FALSE, eval = TRUE}
table(HAM, ILL)
```

We can manipulate the output directly but it is easier if we instruct `R` to save the output of the `table()` function in an object:

```{r, eval = TRUE}
tab <- table(HAM, ILL)
```

The `tab` object contains the output of the `table()` function:

```{r, eval = TRUE}
tab
```

As it is stored in an object we can examine its contents on an item by item basis.

The `tab` object is an object of class `table`:

```{r, eval = TRUE}
class(tab)
```

We can extract data from a table object by using indices or row and column co-ordinates:

```{r, eval = TRUE}
tab[1,1]
tab[1,2]
tab[2,1]
```

The numbers in the square brackets refer to the ***position*** (as row and column co-ordinates) of the data item in the table **_not_** the ***values*** of the variables. We can extract data using the values of the row and column variables by enclosing the index values in double quotes ("). For example:

```{r, eval = TRUE}
tab["1","1"]
```

The two methods of extracting data may be combined. For example:

```{r, eval = TRUE}
tab[1,"1"]
```

We can calculate a risk ratio using the extracted data:

```{r, eval = FALSE}
(tab[1,1]/(tab[1,1]+tab[1,2]))/(tab[2,1]/(tab[2,1]+tab[2,2]))
```

Which returns a risk ratio of

```{r, echo = FALSE, eval = TRUE}
2.044444
```

This is a tedious calculation to have to type in every time you need to calculate a risk ratio from a two-by-two table. It would be better to have a function that calculates and displays the risk ratio automatically. Fortunately, `R` allows us to do just that.
