# Manipulating objects and creating new functions {#exercise2}

In this exercise we will explore how to manipulate `R` objects and how to write functions that can manipulate and extract data and information from `R` objects and produce useful analyses.

Before we go any further we should start `R` and retrieve a dataset:

```{r, eval = TRUE}
salex <- read.table("salex.dat", header = TRUE, na.strings = "9")
```

Missing values are coded as 9 throughout this dataset so we can use the `na.strings` parameter of the `read.table()` function to replace all 9’s with the special `NA` code when we retrieve the dataset. Check that this works by examining the `salex` data.frame:

```{r, eval = TRUE}
salex
names(salex)
```

This data comes from a food-borne outbreak. On Saturday 17th October 1992, eighty-two people attended a buffet meal at a sports club. Within fourteen to twenty-four hours, fifty-one of the participants developed diarrhoea, with nausea, vomiting, abdominal pain and fever.

The columns in the dataset are as follows:

+---------------+----------------------+
| **ILL**       | Ill or not-ill       |
+---------------+----------------------+
| **HAM**       | Baked ham            |
+---------------+----------------------+
| **BEEF**      | Roast beef           |
+---------------+----------------------+
| **EGGS**      | Eggs                 |
+---------------+----------------------+
| **MUSHROOM**  | Mushroom flan        |
+---------------+----------------------+
| **PEPPER**    | Pepper flan          |
+---------------+----------------------+
| **PORKPIE**   | Pork pie             |
+---------------+----------------------+
| **PASTA**     | Pasta salad          |
+---------------+----------------------+
| **RICE**      | Rice salad           |
+---------------+----------------------+
| **LETTUCE**   | Lettuce              |
+---------------+----------------------+
| **TOMATO**    | Tomato salad         |
+---------------+----------------------+
| **COLESLAW**  | Coleslaw             |
+---------------+----------------------+
| **CRISPS**    | Crisps               |
+---------------+----------------------+
| **PEACHCAKE** | Peach cake           |
+---------------+----------------------+
| **CHOCOLATE** | Chocolate cake       |
+---------------+----------------------+
| **FRUIT**     | Tropical fruit salad |
+---------------+----------------------+
| **TRIFLE**    | Trifle               |
+---------------+----------------------+
| **ALMONDS**   | Almonds              |
+---------------+----------------------+

Data is available for seventy-seven of the eighty-two people who attended the sports club buffet. All of the variables are coded 1=yes, 2=no.

We can use the `attach()` function to make it easier to access our data: 

```{r, eval = TRUE}
attach(salex)
```

The two-by-two table is a basic epidemiological tool. In analysing data from a food-borne outbreak collected as a retrospective cohort study, for example, we would tabulate each exposure (suspect foodstuffs) against the outcome (illness) and calculate risk ratios and confidence intervals. `R` has no explicit function to calculate risk ratios from two-by-two tables but we can easily write one ourselves.

The first step in writing such a function would be to create the two-by-two table. This can be done with the `table()` function. We will use a table of `HAM` by `ILL` as an illustration:

```{r, eval = FALSE}
table(HAM, ILL)
```

This command produces the following output:

```{r, echo = FALSE, eval = TRUE}
table(HAM, ILL)
```

We can manipulate the output directly but it is easier if we instruct `R` to save the output of the `table()` function in an object:

```{r, eval = TRUE}
tab <- table(HAM, ILL)
```

The `tab` object contains the output of the `table()` function:

```{r, eval = TRUE}
tab
```

As it is stored in an object we can examine its contents on an item by item basis.

The `tab` object is an object of class `table`:

```{r, eval = TRUE}
class(tab)
```

We can extract data from a table object by using indices or row and column co-ordinates:

```{r, eval = TRUE}
tab[1,1]
tab[1,2]
tab[2,1]
```

The numbers in the square brackets refer to the ***position*** (as row and column co-ordinates) of the data item in the table **_not_** the ***values*** of the variables. We can extract data using the values of the row and column variables by enclosing the index values in double quotes ("). For example:

```{r, eval = TRUE}
tab["1","1"]
```

The two methods of extracting data may be combined. For example:

```{r, eval = TRUE}
tab[1,"1"]
```

We can calculate a risk ratio using the extracted data:

```{r, eval = FALSE}
(tab[1,1]/(tab[1,1]+tab[1,2]))/(tab[2,1]/(tab[2,1]+tab[2,2]))
```

Which returns a risk ratio of

```{r, echo = FALSE, eval = TRUE}
2.044444
```

This is a tedious calculation to have to type in every time you need to calculate a risk ratio from a two-by-two table. It would be better to have a function that calculates and displays the risk ratio automatically. Fortunately, `R` allows us to do just that.

The `function()` function allows us to create new functions in `R`:

```{r, eval = FALSE}
tab2by2 <- function(exposure, outcome) {}
```

This creates an empty function called `tab2by2` that expects two parameters called `exposure` and `outcome`. We could type the whole function in at the `R` command prompt but it is easier to use a text editor:

```{r, eval = FALSE}
fix(tab2by2)
```

This will start an editor with the empty `tab2by2()` function already loaded. We can now edit this function to make it do something useful:

```{r, eval = FALSE}
function(exposure, outcome)
  {
  tab <- table(exposure, outcome)
  a <- tab[1,1]
  b <- tab[1,2]
  c <- tab[2,1]
  d <- tab[2,2]
  rr <- (a / (a + b)) / (c / (c + d))
  print(tab)
  print(rr) 
  }
```

```{r, echo = FALSE, eval = TRUE}
tab2by2 <- function(exposure, outcome) {
  tab <- table(exposure, outcome)
  a <- tab[1,1]
  b <- tab[1,2]
  c <- tab[2,1]
  d <- tab[2,2]
  rr <- (a / (a + b)) / (c / (c + d))
  print(tab)
  print(rr) 
}
```

Once you have made the changes shown above, check your work, save the file, and quit the editor.
Before proceeding we should examine the `tab2by2()` function to make sure we understand what the function will do:

* The first line defines `tab2by2` as a function that expects to be given two parameters which are called `exposure` and `outcome`.

* The body of the function (i.e. the work of the function) is enclosed within curly brackets (`{}`).

* The first line of the body of the function creates a table object (`tab`) using the variables specified when the `tab2by2()` function is called (these are the parameters `exposure` and `outcome`).

* The next line creates four new objects (called `a`, `b`, `c`, and `d`) which contain the values of the four cells in the two-by-two table.

* The following line calculates the risk ratio using the objects `a`, `b`, `c`, and `d` and stores the result of the calculation in an object called `rr`.

* The final two lines print the contents of the `tab` and `rr` objects. 

Let's try the `tab2by2()` function with our test data:

```{r, eval = TRUE}
tab2by2(HAM, ILL)
```

The `tab2by2()` function displays a table of `HAM` by `ILL` followed by the risk ratio calculated from the data in the table.

Try producing another table:

```{r, eval = TRUE}
tab2by2(PASTA, ILL)
```

Have a look at the `R` objects available to you: 

```{r, eval = TRUE}
ls()
```

Note that there are no `a`, `b`, `c`, `d`, or `rr` objects. 

Examine the `tab` object:

```{r, eval = TRUE}
tab
```

This is the table of `HAM` by `ILL` that you created earlier **_not_** the table of `PASTA` by `ILL` that was created by the `tab2by2()` function.

The `tab`, `a`, `b`, `c`, `d`, and `rr` objects in the `tab2by2()` function are local to that function and do not change anything outside of that function. This means that the `tab` object inside the function is independent of any object of the same name outside of the function.

When a function completes its work, all of the objects that are local to that function are automatically removed. This is useful as it means that you can use object names inside functions that will not interfere with objects of the same name that are stored elsewhere. It also means that you do not clutter up the `R` workspace with temporary objects.

Just to prove that `tab` in the `tab2by2()` function exists only in the `tab2by2()` function we can delete the tab object from the `R` workspace:

```{r, eval = TRUE}
rm(tab)
```

Now try another call to the `tab2by2()` function: 

```{r, eval = TRUE}
tab2by2(FRUIT, ILL)
```

Now list the `R` objects available to you: 

```{r, eval = TRUE}
ls()
```

Note that there are no `tab`, `a`, `b`, `c`, `d`, or `rr` objects.

The `tab2by2()` function is very limited. It only displays a table and calculates and displays a simple ratio. A more useful function would also calculate and display a confidence interval for the risk ratio. This is what we will do now. Use the `fix()` function to edit the `tab2by2()` function:

```{r, eval = FALSE}
fix(tab2by2)
```

