# Extending R with packages {#exercise5}

`R` has no built-in functions for survival analysis but, because it is an extensible system, survival analysis is
available as an add-in package. You can find a list of add-in packages at the `R` website. 

http://www.r-project.org/

Add-in packages are installed from the Internet. There are a series of `R` functions that enable you to download and install add-in packages.

The `survival` package adds functions to `R` that enable it to analyse survival data. This package may be downloaded and installed using `install.packages("survival")` or from the `Packages` or `Packages & Data` menu if you are using a GUI version of `R`.

Packages are loaded into `R` as they are needed using the `library()` function. Start `R` and load the `survival` package:

&nbsp;

```{r, eval = TRUE}
library(survival)
```

&nbsp;

Before we go any further we should retrieve a dataset:

&nbsp;

```{r, eval = TRUE}
ca <- read.table("ca.dat", header = TRUE)
attach(ca)
```

&nbsp;

The columns in this dataset on the survival of cancer patients in two different treatment groups are as follows:

&nbsp;

+---------------+----------------------------------------+
| **time**      | Survival or censoring time (months)    |
+---------------+----------------------------------------+
| **status**    | Censoring status (1=dead, 0=censored)  |
+---------------+----------------------------------------+
| **group**     | Treatment group (1 / 2)                |
+---------------+----------------------------------------+

&nbsp;

We next need to create a `survival` object from the `time` and `status` variables using the `Surv()` function: 

&nbsp;

```{r, eval = TRUE}
response <- Surv(time, status)
```

&nbsp;

We can then specify the model for the survival analysis. In this case we state that survival (`response`) is dependent upon the treatment `group`:

&nbsp;

```{r, eval = TRUE}
ca.surv <- survfit(response ~ group)
```

&nbsp;

The `summary()` function applied to a `survfit` object lists the survival probabilities at each time point with 95% confidence intervals:

&nbsp;

```{r, eval = TRUE}
summary(ca.surv)
```

&nbsp;

Printing the `ca.surv` object provides another view of the results: 

&nbsp;

```{r, eval = TRUE}
ca.surv
```

&nbsp;

The `plot()` function with a `survfit` object displays the survival curves: 

&nbsp;

```{r, eval = TRUE, fig.align = "center", fig.pos = "H", fig.retina = 1}
plot(ca.surv, xlab = "Months", ylab = "Survival")
```

&nbsp;

We can make it easier to distinguish between the two lines by specifying a width for each line using thelwd
parameter of the `plot()` function:

&nbsp;

```{r, eval = TRUE, fig.align = "center", fig.pos = "H", fig.retina = 1}
plot(ca.surv, xlab = "Months", ylab = "Survival", lwd = c(1, 2))
```

&nbsp;

It would also be useful to add a legend:

&nbsp;

```{r, eval = FALSE}
legend(125, 1, names(ca.surv$strata), lwd = c(1, 2))
```

```{r, echo = FALSE, eval = TRUE, fig.align = "center", fig.pos = "H", fig.retina = 1}
plot(ca.surv, xlab = "Months", ylab = "Survival", lwd = c(1, 2))
legend(125, 1, names(ca.surv$strata), lwd = c(1, 2))
```

&nbsp;

If there is only one survival curve to plot then plotting a `survfit` object will plot the survival curve with 95% confidence limits. You can specify that confidence limits should be plotted when there is more than one survival curve but the results can be disappointing:

&nbsp;

```{r, eval = TRUE, fig.align = "center", fig.pos = "H", fig.retina = 1}
plot(ca.surv, conf.int = TRUE)
```

&nbsp;

Plots can be improved by specifying different colours for each curve:

&nbsp;

```{r, eval = TRUE, fig.align = "center", fig.pos = "H", fig.retina = 1}
plot(ca.surv, conf.int = TRUE, col = c("red", "darkgreen"))
```

&nbsp;

We can perform a formal test of the two survival times using the `survdiff()` function: 

&nbsp;

```{r, eval = TRUE}
survdiff(response ~ group)
```

&nbsp;

We can now quit `R`: 

&nbsp;

```{r, eval = FALSE}
q()
```

&nbsp;

For this exercise there is no need to save the workspace image so click the **No** or **Don't Save** button (GUI) or enter `n` when prompted to save the workspace image (terminal).

## Summary

* `R` can be extended by adding additional packages. Some packages are included with the standard `R` installation but many others are available and may be downloaded from the Internet.

* You can find a list of add-in packages at the `R` website: http://www.r-project.org/

* Packages may also be downloaded and installed from this site using the `install.packages()` function or from the **Packages** or **Packages & Data** menu if you are using a GUI version of `R`.

* Packages are loaded into `R` as they are needed using the `library()` function. You can use the `search()` function to display a list of loaded packages and attached data.frames.
