# Making your own objects behave like R objects {#exercise6}

In the previous exercises we concentrated on writing functions that take some input data, analyse it, and display the results of the analysis. The standard `R` functions we have used all do this. The `fisher.test()` function, for example, takes a `table` object (or the names of two variables) as input and calculates and displays the p- value for *Fisher's exact test* and the odds ratio and associated confidence interval for two-by-two tables:

&nbsp;

```{r, eval = TRUE}
fem <- read.table("fem.dat", header = TRUE)
attach(fem)
fisher.test(SEX, LIFE)
```

&nbsp;

The results of the `fisher.test()` function may also be saved for later use: 

&nbsp;

```{r, eval = TRUE}
ft <- fisher.test(SEX, LIFE)
ft
```

&nbsp;

The `fisher.test()` function returns an object of the class `htest`: 

&nbsp;

```{r, eval = TRUE}
class(ft)
```

&nbsp;

which is a list containing the output of the `fisher.test()` function. Each item of output is stored as a different named item in the list:

&nbsp;

```{r, eval = FALSE}
names(ft)
str(ft)
```

```{r, echo = FALSE, eval = TRUE}
names(ft)
str(ft)
```

&nbsp;

Each of these items can be referred to by name:

&nbsp;

```{r, eval = FALSE}
ft$estimate
ft$conf.int
```

```{r, echo = FALSE, eval = TRUE}
ft$estimate
ft$conf.int
```

&nbsp;

When you display the output of the `fisher.test()` function either by calling the function directly: 

&nbsp;

```{r, eval = TRUE}
fisher.test(SEX, LIFE)
```

&nbsp;

or by typing the name of an object created using the `fisher.test()` function: 

&nbsp;

```{r, eval = TRUE}
ft
```

&nbsp;

The `print()` function takes over and formatted output is produced. The `print()` function knows about `htest` class objects and produces output of the correct format for that class of object. This means that any function that produces an `htest` object (or any other standard `R` object) does not need to include `R` commands to produce formatted output.

All hypothesis testing functions supplied with `R` produce objects of the htest class and use the `print()` function to produce formatted output. For example:

&nbsp;

```{r, eval = FALSE}
tt <- t.test(WT ~ LIFE)
class(tt)
tt
```

```{r, echo = FALSE, eval = TRUE}
tt <- t.test(WT ~ LIFE)
class(tt)
tt
```

&nbsp;

You can use this feature of `R` in your own functions. We will explore this by writing a function to test the null hypothesis that the *variance to mean ratio* of a vector of numbers is equal to one. Such a test might be used to investigate the spatial distribution (e.g. over natural sampling units such as households) of cases of a disease.

Create a new function using the `function()` function:

&nbsp;

```{r, eval = FALSE}
v2m.test <- function(data) {}
```

&nbsp;

And start the function editor:

&nbsp;

```{r, eval = FALSE}
fix(v2m.test)
```

\newpage

Now edit this function to make it do something useful:

&nbsp;

```{r, eval = FALSE}
function(data) {
  nsu <- length(data)
  obs <- sum(data)
  m <- obs / nsu
  v <- var(data)
  vmr <- v / m
  chi2 <- sum((data - m)^2) / m
  df <- nsu - 1
  p <- 1 - pchisq(chi2, df)
  names(chi2) <- "Chi-square"
  names(df) <- "df"
  names(vmr) <- "Variance : mean ratio"
  v2m <- list(method = "Variance to mean test",
              data.name = deparse(substitute(data)),
              statistic = chi2,
              parameter = df,
              p.value = p,
              estimate = vmr)
  class(v2m) <- "htest"
  return(v2m)
}
```

```{r, echo = FALSE, eval = TRUE}
v2m.test <- function(data) {
  nsu <- length(data)
  obs <- sum(data)
  m <- obs / nsu
  v <- var(data)
  vmr <- v / m
  chi2 <- sum((data - m)^2) / m
  df <- nsu - 1
  p <- 1 - pchisq(chi2, df)
  names(chi2) <- "Chi-square"
  names(df) <- "df"
  names(vmr) <- "Variance : mean ratio"
  v2m <- list(method = "Variance to mean test",
              data.name = deparse(substitute(data)),
              statistic = chi2,
              parameter = df,
              p.value = p,
              estimate = vmr)
  class(v2m) <- "htest"
  return(v2m)
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file, and quit the editor.

Before proceeding we should examine the `v2m.test()` function to make sure we understand what is happening:
  
1. The first eight lines after the opening curly bracket (`{`) contain the required calculations.

2. The next three lines use the `names()` function to give our variables names that will make sense in formatted output.

3. The next line creates a list of items that the function returns using some of the names used by `htest` class objects.

4. The next line tells `R` that the list object called `v2m` is of the class `htest`.

5. The next line causes the function to return the `v2m` object (i.e. a list of class `htest` containing the named items `method`, `data.name`, `statistic`, `parameter`, `p.value`, and `estimate`). 

6. The final line ends the function definition.
    
Note that objects of class htest may contain items with the following names:

&nbsp;
 
+--------------------+-----------------------------------------------------------------------+
| **Item**           | **Usage**                                                             |
+====================+=======================================================================+
| **method**         | Text description of the test used to title output                     |
+--------------------+-----------------------------------------------------------------------+
| **data.name**      | Name(s) of data or variables used for the test                        |
+--------------------+-----------------------------------------------------------------------+
| **null.value**     | The null value                                                        |
+--------------------+-----------------------------------------------------------------------+
| **statistic**      | Value of test statistic                                               |
+--------------------+-----------------------------------------------------------------------+
| **parameter**      | A test parameter such as the degrees of freedom of the test statistic |
+--------------------+-----------------------------------------------------------------------+
| **p.value**        | The p-value of the test                                               |
+--------------------+-----------------------------------------------------------------------+
| **estimate**       | An estimate (e.g. the mean)                                           |
+--------------------+-----------------------------------------------------------------------+
| **conf.int**       | Confidence interval of estimate                                       |
+--------------------+-----------------------------------------------------------------------+
| **alternative**    | Text describing the alternative hypothesis                            |
+--------------------+-----------------------------------------------------------------------+
| **note**           | Text note                                                             |
+--------------------+-----------------------------------------------------------------------+

&nbsp;

We are now ready to test the `v2m.test()` function. This table: 

&nbsp;

```
Number of cases :       0  1  2  3  4  6
Number of households : 24 29 26 14  5  2
```

&nbsp;

shows the number of cases of chronic (stunting) undernutrition found in a random sample of 100 households. 

We can reproduce the data behind this table using a combination of the `c()` and `rep()` functions:

&nbsp;

```{r, eval = TRUE}
stunt <- c(rep(0,24), rep(1,29), rep(2,26), rep(3,14), rep(4,5),
           rep(5,0), rep(6,2))
table(stunt)
```

&nbsp;

And use it to test our new `v2m.test()` function: 

&nbsp;

```{r, eval = FALSE}
v2m.test(stunt)
```

&nbsp;

Which should produce the following output:

&nbsp;

```{r, eval = TRUE}
v2m.test(stunt)
```

&nbsp;

If your `vm2.test()` function does not produce this output then use the `fix()` function: 

&nbsp;

```{r, eval = FALSE}
fix(v2m.test)
```

&nbsp;

to check and edit the `vm2.test()` function and try again.

The important thing to note from this exercise is that `R` allows us to specify a class for the output of our functions. This means that we can use standard `R` classes and functions to (e.g.) produce formatted output without us having to write commands to format the output ourselves.

More importantly, it also means that we can write functions that return values when we need them to return values but can also produce formatted output when we need them to produce formatted output.

Our `v2m.test()` function can produce values for later use: 

&nbsp;

```{r, eval = TRUE}
vm <- v2m.test(stunt)
vm$p.value
```

&nbsp;

or produce formatted output:

&nbsp;

```{r, eval = TRUE}
v2m.test(stunt)
```

&nbsp;

This way of working is not limited to using standard `R` classes and functions.

`R` also allows us to define our own classes. We will explore this by defining functions and a new class to deal
with two-by-two tables.

We need to create two functions:

1. One function will handle the calculations.

2. A second function function will produce formatted output when required. 

Create a new function using the `function()` function:

&nbsp;

```{r, eval = FALSE}
rr22 <- function(exposure, outcome) {}
```

&nbsp;

And start the function editor:

&nbsp;

```{r, eval = FALSE}
fix(rr22)
```

&nbsp;

Now edit this function to make it do something useful:

&nbsp;

```{r, eval = FALSE}
function(exposure, outcome) {
  tab <- table(exposure, outcome)
  a <- tab[1,1]
  b <- tab[1,2]
  c <- tab[2,1]
  d <- tab[2,2]
  rr <- (a / (a + b)) / (c / (c + d))
  se.log.rr <- sqrt((b / a) / (a + b) + (d / c) / (c + d))
  lci <- exp(log(rr) - 1.96 * se.log.rr)
  uci <- exp(log(rr) + 1.96 * se.log.rr)
  rr22.output <- list(estimate = rr, ci = c(lci, uci))
  class(rr22.output) <- "rr22"
  return(rr22.output)
}
```

```{r, echo = FALSE, eval = TRUE}
rr22 <- function(exposure, outcome) {
  tab <- table(exposure, outcome)
  a <- tab[1,1]
  b <- tab[1,2]
  c <- tab[2,1]
  d <- tab[2,2]
  rr <- (a / (a + b)) / (c / (c + d))
  se.log.rr <- sqrt((b / a) / (a + b) + (d / c) / (c + d))
  lci <- exp(log(rr) - 1.96 * se.log.rr)
  uci <- exp(log(rr) + 1.96 * se.log.rr)
  rr22.output <- list(estimate = rr, ci = c(lci, uci))
  class(rr22.output) <- "rr22"
  return(rr22.output)
}
```

&nbsp;

Once you have made the changes shown above, save the file and quit the editor.

The `rr22()` function is similar to the `tab2by2()` function that you created in the second exercise of this
tutorial except that the function now returns a list of values instead of formatted output:

&nbsp;

```{r, eval = FALSE}
fem <- read.table("fem.dat", header = TRUE)
attach(fem)
rr22.test <- rr22(SEX, LIFE)
names(rr22.test)
rr22.test$estimate
rr22.test$conf.int
rr22.test$conf.int[1]
rr22.test$conf.int[2]
```

```{r, echo = FALSE, eval = TRUE}
fem <- read.table("fem.dat", header = TRUE)
attach(fem)
rr22.test <- rr22(SEX, LIFE)
names(rr22.test)
rr22.test$estimate
rr22.test$conf.int
rr22.test$conf.int[1]
rr22.test$conf.int[2]
```

&nbsp;

The function returns a list of class `rr22`: 

&nbsp;

```{r, eval = TRUE}
class(rr22.test)
```

&nbsp;

The displayed output from the `rr22()` function is, however, not pretty: 

&nbsp;

```{r, eval = FALSE}
print(rr22.test)
rr22(SEX, LIFE)
```

```{r, echo = FALSE, eval = TRUE}
print(rr22.test)
rr22(SEX, LIFE)
```

&nbsp;

This can be fixed by creating a new function:

&nbsp;

```{r, eval = FALSE}
print.rr22 <- function(x) {}
```

&nbsp;

And start the function editor:

&nbsp;

```{r, eval = FALSE}
fix(print.rr22)
```

&nbsp;

Now edit this function to make it do something useful:

&nbsp;

```{r, eval = FALSE}
function(x) {
  cat("RR     : ", x$estimate, "\n",
      "95% CI : ", x$ci[1], "; ", x$ci[2], "\n", sep = "")
}
```

```{r, echo = FALSE, eval = TRUE}
print.rr22 <- function(x) {
  cat("RR     : ", x$estimate, "\n",
      "95% CI : ", x$ci[1], "; ", x$ci[2], "\n", sep = "")
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file, and quit the editor.

The function name `print.rr22()` indicates that this function contains the print method for objects of class `rr22`. All objects of class `rr22` will use the function `print.rr22()` instead of the standard `R` `print()` function to produce formatted output:

&nbsp;

```{r, eval = FALSE}
rr22(SEX, LIFE)
rr22.test <- rr22(SEX, LIFE)
rr22.test
print(rr22.test)
```

```{r, echo = FALSE, eval = TRUE}
rr22(SEX, LIFE)
rr22.test <- rr22(SEX, LIFE)
rr22.test
print(rr22.test)
```

&nbsp;

Note that we can still extract returned values from an `rr22` class object: 

&nbsp;

```{r, eval = FALSE}
rr22.test$estimate
```

&nbsp;

The `print.rr22()` function only controls the way an entire `rr22` object is displayed.

You might like to use the `save()` function to save the `v2m.test()`, `rr22()`, and `print.rr22()` functions before quitting `R`. We can now quit `R`:

&nbsp;

```{r, eval = FALSE}
q()
```

&nbsp;

For this exercise there is no need to save the workspace image so click the **No** or **Don't Save** button (GUI) or enter `n` when prompted to save the workspace image (terminal).

## Summary

* `R` objects can be assigned a class or type.

* Objects of a specific class or type may share functions that extract and manipulate data common to members of that class. This allows you to write functions that handle data that is common to all members of that class (e.g. to produce formatted output for hypothesis testing functions).

* `R` provides a set of ready-made classes (e.g. `htest`) which can be used by standard R functions such as the `print()` and `summary()` functions.

* `R` allows you to create new classes and class-specific functions that can extract and manipulate data common to the new classes.

* Classes allows you to create versatile functions that return values when you need them to return values but can also produce formatted output when you need them to produce formatted output.

* Classes allow you to write functions that can be chained together so that the output of one function is the input of another function.
