# Writing your own graphical functions {#exercise7}

`R` provides a pretty full set of graphical functions for plotting data as well as `plot()` methods for a wide variety of statistical functions. There will be times, however, when you will need to write you own graphical functions to present and analyse data in a specific way. In this exercise we will create a function that produces a plot that may be used for assessing agreement between two methods of clinical measurement as described in:

```
Bland JM, Altman DG. Statistical Methods for Assessing Agreement Between Two Methods of Clinical Measurement. Lancet. 1986;1: 307–310. 
```

Which involves plotting the difference of two measurements against the mean of the two measurements and calculating and displaying limits of agreement.

Start `R` and retrieve and attach the sample dataset:

```{r, eval = TRUE}
ba <- read.table("ba.dat", header = TRUE)
attach(ba)
```

The `ba` data.frame contains measurements (in litres per minute) taken with a *Wright peak flow meter* and a *Mini-Wright peak flow meter*. This is the same data that is presented in the referenced Lancet article:

```{r, echo = FALSE, eval = TRUE}
ba
```

   Wright Mini
 1 494    512
 2 395    430
 3 516    520
 4 434    428
 5 476    500
 6 557    600
 7 413    364
 8 442    380
 9 650    658
10 433    445
11 417    432
12 656    626
13 267    260
14 478    477
15 178    259
16 423    350
17 427    451
   
You can examine the `ba` data.frame using the `print()` and `summary()` functions:

```{r, eval = FALSE}
print(ba)
ba
summary(ba)
```

```{r, echo = FALSE, eval = TRUE}
print(ba)
ba
summary(ba)
```

The `function()` function allows us to create new functions in `R`: 

```{r, eval = FALSE}
ba.plot <- function(a, b) {}
```

This creates an empty function called `ba.plot()` that expects two parameters called `a` and `b`. We could type the whole function in at the `R` command prompt but it is easier to use a text editor:

```{r, eval = FALSE}
fix(ba.plot)
```

We will start be writing a basic function which we will gradually improve throughout this exercise.

