# Writing your own graphical functions {#exercise7}

`R` provides a pretty full set of graphical functions for plotting data as well as `plot()` methods for a wide variety of statistical functions. There will be times, however, when you will need to write you own graphical functions to present and analyse data in a specific way. In this exercise we will create a function that produces a plot that may be used for assessing agreement between two methods of clinical measurement as described in:

```
Bland JM, Altman DG. Statistical Methods for Assessing Agreement Between Two Methods 
    of Clinical Measurement. The Lancet. 1986;1: 307–310. 
```

Which involves plotting the difference of two measurements against the mean of the two measurements and calculating and displaying limits of agreement.

Start `R` and retrieve and attach the sample dataset:

&nbsp;

```{r, eval = TRUE}
ba <- read.table("ba.dat", header = TRUE)
attach(ba)
```

&nbsp;

The `ba` data.frame contains measurements (in litres per minute) taken with a *Wright peak flow meter* and a *Mini-Wright peak flow meter*. This is the same data that is presented in the referenced Lancet article:

&nbsp;

```{r, echo = FALSE, eval = TRUE}
ba
```

&nbsp;

You can examine the `ba` data.frame using the `print()` and `summary()` functions:

&nbsp;

```{r, eval = FALSE}
print(ba)
ba
summary(ba)
```

```{r, echo = FALSE, eval = TRUE}
print(ba)
ba
summary(ba)
```

&nbsp;

The `function()` function allows us to create new functions in `R`: 

&nbsp;

```{r, eval = FALSE}
ba.plot <- function(a, b) {}
```

&nbsp;

This creates an empty function called `ba.plot()` that expects two parameters called `a` and `b`. We could type the whole function in at the `R` command prompt but it is easier to use a text editor:

&nbsp;

```{r, eval = FALSE}
fix(ba.plot)
```

&nbsp;

We will start be writing a basic function which we will gradually improve throughout this exercise.

Edit the `ba.plot()` function to read:

&nbsp;

```{r, eval = FALSE}
function(a, b) {
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two)
}
```

```{r, echo = FALSE, eval = TRUE}
ba.plot <- function(a, b) {
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two)
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file, and quit the editor.

The function calculates the mean and the difference of the two measures and then plots the results. Lets try the
`ba.plot()` function with the test data: 

&nbsp;

```{r, eval = TRUE}
ba.plot(Wright, Mini)
```

&nbsp;

The resulting plot is rather plain and lacks meaningful titles and axis labels. Use the `fix()` function to edit the `ba.plot()` function:

&nbsp;

```{r, eval = FALSE}
fix(ba.plot)
```

&nbsp;

Edit the function to read:

&nbsp;

```{r, eval = FALSE}
function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title)
}
```

```{r, echo = FALSE, eval = TRUE}
ba.plot <- function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title)
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file, and quit the editor.

We have added a new parameter (`title`) to the function and given this a default value of `Bland and Altman Plot`. Adding `title` as a parameter means that we will be able to specify a title for the plot when we call the function. We have also used the function combination `deparse(substitute())` to retrieve the names of the vectors passed to parameters `a` and `b`. The `paste()` function pastes pieces of text together. It is used here to create the text for the axis labels used with the `plot()` function.

Lets try the `ba.plot()` function with the test data: 

&nbsp;

```{r, eval = TRUE}
ba.plot(Wright, Mini)
```

&nbsp;

We may also specify a title for the plot using the title parameter:

&nbsp;

```{r, eval = TRUE}   
ba.plot(Wright, Mini, title = "PEFR data")
```

&nbsp;

We can now edit the function to calculate and plot mean, difference, and the limits of agreement. Use the `fix()` function to edit the `ba.plot()` function:

&nbsp;

```{r, eval = FALSE}
fix(ba.plot)
```

&nbsp;

Edit the function to read:

&nbsp;

```{r, eval = FALSE}
function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title) 
  mean.diff <- mean(diff.two)
  sd.diff <- sd(diff.two)
  upper <- mean.diff + 1.96 * sd.diff
  lower <- mean.diff - 1.96 * sd.diff
  lines(x = range(mean.two), y = c(mean.diff, mean.diff), lty = 3) 
  lines(x = range(mean.two), y = c(upper, upper), lty = 3)
  lines(x = range(mean.two), y = c(lower, lower), lty = 3)
}
```

```{r, echo = FALSE, eval = TRUE}
ba.plot <- function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title) 
  mean.diff <- mean(diff.two)
  sd.diff <- sd(diff.two)
  upper <- mean.diff + 1.96 * sd.diff
  lower <- mean.diff - 1.96 * sd.diff
  lines(x = range(mean.two), y = c(mean.diff, mean.diff), lty = 3) 
  lines(x = range(mean.two), y = c(upper, upper), lty = 3)
  lines(x = range(mean.two), y = c(lower, lower), lty = 3)
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file and quit the editor.

We have used the `mean()` and `sd()` functions to calculate the mean and standard deviation of the difference between the two measures and calculated the limits of agreement (`upper` and `lower`) assuming that the differences are *Normally* distributed.

The `lines()` function is then used to plot the mean and the limits of agreement on top of the existing scatter plot.

The parameter `lty = 3` used with the `lines()` function specifies dotted lines.

`R` provided a great number of graphical parameters that can be used to customise plots. You can see a list of
these parameters using:

&nbsp;

```{r, eval = TRUE}
help(par)
```

&nbsp;

These parameters can be specified for almost all graphical functions.

Lets try the `ba.plot()` function with the test data:

&nbsp;

```{r, eval = TRUE}
ba.plot(Wright, Mini, title = "Difference vs. mean for PEFR data")
```

&nbsp;

The function is almost complete. All that remains to do is to label the lines with the values of the mean difference and the limits of agreement.

Use the `fix()` function to edit the `ba.plot()` function: 

&nbsp;

```{r, eval = FALSE}
fix(ba.plot)
```

&nbsp;

Edit the function to read:

&nbsp;

```{r, eval = FALSE}
function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title) 
  mean.diff <- mean(diff.two)
  sd.diff <- sd(diff.two)
  upper <- mean.diff + 1.96 * sd.diff
  lower <- mean.diff - 1.96 * sd.diff
  lines(x = range(mean.two), y = c(mean.diff, mean.diff), lty = 3) 
  lines(x = range(mean.two), y = c(upper, upper), lty = 3)
  lines(x = range(mean.two), y = c(lower, lower), lty = 3)
  m.text <- round(mean.diff, digits = 1)
  u.text <- round(upper , digits = 1)
  l.text <- round(lower, digits = 1)
  text(max(mean.two), mean.diff, m.text, adj = c(1,1)) 
  text(max(mean.two), upper, u.text, adj = c(1,1)) 
  text(max(mean.two), lower, l.text, adj = c(1,1))
}
```

```{r, echo = FALSE, eval = TRUE}
ba.plot <- function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title) 
  mean.diff <- mean(diff.two)
  sd.diff <- sd(diff.two)
  upper <- mean.diff + 1.96 * sd.diff
  lower <- mean.diff - 1.96 * sd.diff
  lines(x = range(mean.two), y = c(mean.diff, mean.diff), lty = 3) 
  lines(x = range(mean.two), y = c(upper, upper), lty = 3)
  lines(x = range(mean.two), y = c(lower, lower), lty = 3)
  m.text <- round(mean.diff, digits = 1)
  u.text <- round(upper , digits = 1)
  l.text <- round(lower, digits = 1)
  text(max(mean.two), mean.diff, m.text, adj = c(1,1)) 
  text(max(mean.two), upper, u.text, adj = c(1,1)) 
  text(max(mean.two), lower, l.text, adj = c(1,1))
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file, and quit the editor.
We have used the `round()` function to limit the display of the mean difference and the limits of agreement to one decimal place and used the `text()` function to display these (rounded) values. The `adj` parameter to the `text()` function controls the position and justification of text.

Let's try the `ba.plot()` function with the test data: 

&nbsp;

```{r, eval = TRUE}
ba.plot(Wright, Mini, title = "PEFR data")
```

&nbsp;

The graphical function is now complete.

One improvement that we could make is for the function to produce a chart and return the values of the mean difference and the limits of agreement.

We would do this in exactly the same way as we would with a non-graphical function. We would return the mean difference and the limits of agreement as members of a list.

We could also specify a class for the returned list and create a class specific `print()` function (or *method*) to produce nicely formatted output.

Use the `fix()` function to edit the `ba.plot()` function: 

&nbsp;

```{r, eval = FALSE}
fix(ba.plot)
```

&nbsp;

Edit the function to read:

&nbsp;

```{r, eval = FALSE}
function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title) 
  mean.diff <- mean(diff.two)
  sd.diff <- sd(diff.two)
  upper <- mean.diff + 1.96 * sd.diff
  lower <- mean.diff - 1.96 * sd.diff
  lines(x = range(mean.two), y = c(mean.diff, mean.diff), lty = 3) 
  lines(x = range(mean.two), y = c(upper, upper), lty = 3)
  lines(x = range(mean.two), y = c(lower, lower), lty = 3)
  m.text <- round(mean.diff, digits = 1)
  u.text <- round(upper , digits = 1)
  l.text <- round(lower, digits = 1)
  text(max(mean.two), mean.diff, m.text, adj = c(1,1)) 
  text(max(mean.two), upper, u.text, adj = c(1,1)) 
  text(max(mean.two), lower, l.text, adj = c(1,1))
  ba <- list(mean = mean.diff, limits = c(lower, upper))
  class(ba) <- "ba"
  return(ba)
}
```

```{r, echo = FALSE, eval = TRUE}
ba.plot <- function(a, b, title = "Bland and Altman Plot") {
  a.txt <- deparse(substitute(a))
  b.txt <- deparse(substitute(b))
  x.lab <- paste("Mean of", a.txt, "and", b.txt)
  y.lab <- paste(a.txt, "-", b.txt)
  mean.two <- (a + b) / 2
  diff.two <- a - b
  plot(mean.two, diff.two, xlab = x.lab, ylab = y.lab, main = title) 
  mean.diff <- mean(diff.two)
  sd.diff <- sd(diff.two)
  upper <- mean.diff + 1.96 * sd.diff
  lower <- mean.diff - 1.96 * sd.diff
  lines(x = range(mean.two), y = c(mean.diff, mean.diff), lty = 3) 
  lines(x = range(mean.two), y = c(upper, upper), lty = 3)
  lines(x = range(mean.two), y = c(lower, lower), lty = 3)
  m.text <- round(mean.diff, digits = 1)
  u.text <- round(upper , digits = 1)
  l.text <- round(lower, digits = 1)
  text(max(mean.two), mean.diff, m.text, adj = c(1,1)) 
  text(max(mean.two), upper, u.text, adj = c(1,1)) 
  text(max(mean.two), lower, l.text, adj = c(1,1))
  ba <- list(mean = mean.diff, limits = c(lower, upper))
  class(ba) <- "ba"
  return(ba)
}
```

&nbsp;

Once you have made the changes shown above, save the file and quit the editor. 

Create a `print()` function for objects of the `ba` class:

&nbsp;

```{r, eval = FALSE}
print.ba <- function(x) {}
```

&nbsp;

Use the `fix()` function to edit the new function: 

&nbsp;

```{r, eval = FALSE}
fix(print.ba)
```

&nbsp;

Edit the function to read:

&nbsp;

```{r, eval = FALSE}
function(x) {
  cat("Mean difference     : ", x$mean, "\n",
      "Limits of agreement : ", x$limits[1], "; ", x$limits[2], "\n",
      sep = "")
}
```

```{r, echo = FALSE, eval = TRUE}
print.ba <- function(x) {
  cat("Mean difference     : ", x$mean, "\n",
      "Limits of agreement : ", x$limits[1], "; ", x$limits[2], "\n",
      sep = "")
}
```

&nbsp;

Once you have made the changes shown above, check your work, save the file, and quit the editor.

Let's try the `ba.plot()` function with the test data: 

&nbsp;

```{r, eval = TRUE}
ba.plot(Wright, Mini, title = "PEFR data")
```

&nbsp;

The function produces the plot and returns the mean difference and limits of agreement as a list of class `ba` which is formatted and printed by the `print.ba()` function.

We can manipulate the returned values just as we would with any other function:

&nbsp;

```{r, eval = FALSE}
ba.test <- ba.plot(Wright, Mini)
print(ba.test)
ba.test
ba.test$mean
ba.test$limits
ba.test$limits[1]
ba.test$limits[2]
```

```{r, echo = FALSE, eval = TRUE}
ba.test <- ba.plot(Wright, Mini)
print(ba.test)
ba.test
ba.test$mean
ba.test$limits
ba.test$limits[1]
ba.test$limits[2]
```

&nbsp;

You might like to use the `save()` function to save the `ba.plot()` and `print.ba()` functions before quitting `R`.

We can now quit `R`: 

&nbsp;

```{r, eval = FALSE}
q()
```

&nbsp;

For this exercise there is no need to save the workspace image so click the **No** or **Don't Save** button (GUI) or enter `n` when prompted to save the workspace image (terminal).

## Summary

* `R` allows you to create functions that produce graphical output.

* `R` allows you to create functions that produce graphical output and return values.

* `R` objects can be assigned a class or type.

* `R` allows you to create new classes and class-specific functions that can extract and manipulate data common to the new classes.

* Classes allows you to create versatile functions that return values when we need them to return values but can also produce formatted output when we need them to produce formatted output.

* Classes allow you to write functions that can be chained together so that the output of one function is the input of another function.
